From physical principles to relativistic classical Hamiltonian and Lagrangian particle mechanics

### 2015-02-06 2014CompleteClassicalMechanics/completeClassicalMechanics.tex
Reviewed second section
Limit definition of states and QT.

Numeric labels are prevalent in physics. For the discrete case, we can use integers and expect a one-to-one map between $z\in\mathbbm{Z}$ and $i\in I$. The deterministic map than becomes $f:\mathbbm{Z} \leftrightarrow \mathbbm{Z}$. For the continuous case, one may simply expect to replace $\mathbbm{Z}$ with $\mathbbm{R}$, but this presents a problem. In the continuous limit, the distribution of our infinitesimal states will become a density which is defined over infinitesimal intervals. Since the distribution and the states must be defined on the same object, the label is really an interval over $\mathbbm{R}$, not a point. This also makes physical sense, as we never deal with points per se, but widths that can be made arbitrarily small. With this in mind, we have the following definitions.

\begin{defn}\label{sdof}
A \emph{degree of freedom} is a state variable identified by an infinitesimal cell of $\mathbbm{R}$.
\end{defn}

@@

\begin{cor}\label{continuousLabels}
Each degree of freedom is identified by the pair of labels $\langle q,k \rangle$. As $k\,dq$ is invariant under coordinate changes, $dk\wedge dq$ is invariant and $k$ is contravariant.
\end{cor}

\begin{prop}\label{continuousLabels}
The cardinality of labels on a degree of freedom is given by $\omega = \hbar \, dq \wedge dk$, where $\hbar$ is a constant that defines the unit for labels.
\end{prop}

Since $q$ and $k$ are independent labels, the number of labels defined over a degree of freedom is the product of the labels defined on $q$ and $k$ (\ref{labelsCombine}). The number of labels for an interval of $q$ or $k$ will be proportional to said interval, therefore we have $\hbar \, dq \wedge dk$, where the constant is given by the choice of units for labels. It is customary to combine $\hbar$ with $k$.

\begin{defn}\label{sdof}
We define the \emph{conjugate momentum} $p=\hbar k$.
\end{defn}

\begin{cor}\label{continuousLabels}
Each degree of freedom is identified by the pair of labels $\langle q,p \rangle$. $\theta_0 = p dq$ and $\omega = dq \wedge dp$ are invariant, $p$ is contravariant.
\end{cor}

\begin{prop}\label{continuousLabelDist}
The configuration state over a degree of freedom is $\mathbbm{c}=\int D(q,p) \mathbbm{s}_{qp} = \int \rho(q,p) dq \wedge dp \mathbbm{s}_{qp}$, where $\rho(q,p)\equiv D(q,p) / \omega$ is the distribution density for each label.
\end{prop}

This is simply the limit of \ref{classicalDistribution} for the continuous case. Note that, since $\mathbbm{c}$ and $\omega$ are invariant under coordinate transformations, so is $\rho$.

In the language of differential geometry, we recognise the function $\rho$, the one-form $\theta_0$ and the two-form $\omega$. They are truly fundamental objects as they are intimately linked to the way states are defined ($\rho$ is the distribution, $\theta_0 / \hbar$ is the cell width) and counted ($\omega$ is the cardinality of the labels) and they don't depend on coordinate choice.

>>>>

The above definitions readily apply for labels identified by integers.\footnote{For each label $i$ there is one and only one integer $z$.} Let $z \in \mathbbm{Z}$, we have $\mathbbm{c}=\sum\limits_{z \in \mathbbm{Z}} D(z) \mathbbm{s}(z)$ and the deterministic map becomes $f:\mathbbm{Z} \leftrightarrow \mathbbm{Z}$. For the continuous case, one may simply expect to replace $z \in \mathbbm{Z}$ with $r \in \mathbbm{R}$, but this does not work. In the continuous limit, we would have $\mathbbm{c}=\int\limits_{r \in \mathbbm{R}} \rho(r) dr \mathbbm{s}(r)$, where $\rho(r) = D(r) / dr$. The continuous distribution $\rho$ is a density, defined over interval $dr$. That is: it's really $\rho(r, dr(r))$, function of both the center and the width of the interval. A bijective map on just $r$ is not sufficient, $dr$ must be mapped as well. On one side we claim the state fully identified by $r$, on the other $dr$ (an different label) is required for the density and the bijective map. We can't have it both ways.

In the continuous case, then, the appropriate label corresponds to a cell over $\mathbbm{R}$, not a point. This also makes physical sense, as we never deal with points per se, but widths that can be made arbitrarily small. An cell corresponds to $\mathbbm{R}^2$: the center and the width. With this in mind, we have the following definitions.

\begin{defn}\label{sdof}
A \emph{degree of freedom} is a state variable identified by an infinitesimal cell of $\mathbbm{R}$.\footnote{For each label $i$ there is one and only infinitesimal cell.}
\end{defn}

@@

\begin{cor}\label{continuousLabels}
Each degree of freedom is identified by the pair of labels $\langle q,k \rangle \in QT^*$. $k\,dq$ is invariant under coordinate changes, $dq\wedge dk$ is invariant and $k$ is contravariant.
\end{cor}

Let's start with the discrete case. Let $Q$ be a finite region of $\mathbbm{R}$. Divide the region in $N$ equal intervals of center $q$ and length $\Delta q$.\footnote{The definition could be mathematically more general. We use equally spaced intervals as it makes the derivation less cumbersome.} Let $K$ be a finite region of $\mathbbm{R}$. Divide the region in $M$ equal intervals of center $k$ and length $\Delta k$. Let $I$ be the label set identified by the cells in $Q$ with center $q(i)$ and width $k(i) \Delta q$. The configuration state will be given by $\mathbbm{c}=\sum\limits_{i \in \mathbbm{I}} \rho(q(i), k(i)) \Delta q \Delta k \mathbbm{s}(q(i), k(i))$, where $\rho(q(i), k(i)) = D(q(i), k(i)) / (\Delta q \Delta k)$.

As we increase $N$ and $M$, the cardinality of the label set $I$ increases, $\Delta q$ and $\Delta k$ decrease and so does the width of the cells. Our definitions, though, do not change: for each $\langle q,k \rangle$ we have one and only one state. In the limit, we will cover every possible center $q$ and evert possible cell number $k$. Our configuration state becomes $\mathbbm{c}=\int\limits_{q \in Q \; k\in K} \rho(q, k) dq \wedge dk \mathbbm{s}(q(i), k(i))$. We can then arbitrarily increase the size of $Q$, and note that none of the definition change.

If we apply a coordinate change $q'=q'(q)$, we change the labels but the state must remain the same, defined at the same point with the same width. $k' dq'$ must be then equal to $k dq$; $k'= k dq / dq'$ is contravariant, $dq' \wedge dk'= dq'/dq dq \wedge dk dq/dq' = dq \wedge dk$ is invariant. This means that $\rho$ is also invariant, which makes sense: the density depends only on the state (the cell), and not the coordinate chosen to represent them. As $k$ is contravariant, in the limit $K$ becomes the cotangent space of $Q$.

\begin{prop}\label{continuousLabels}
The cardinality of a label set $I$, subset of a degree of freedom, is given by $n(I)=\int \omega$, where $\omega = \hbar \, dq \wedge dk$ and $\hbar$ is a constant that defines the unit of state cardinality.
\end{prop}

As we deal with continuous labels, a label set is uncountable: given any range $\delta q$ and $\delta k$ there are an infinite number of possible labels $I$. Let $I_0$ be another set, defined on $\delta q_0$ and $\delta k_0$. In the discrete case, we have:

\begin{align*}
\frac{n(I)}{n(I_0)} = \frac{\delta q \delta k / \Delta q \Delta k}{\delta q_0 \delta k_0 / \Delta q \Delta k} = \frac{\delta q \delta k}{\delta q_0 \delta k_0}
\end{align*}

That is: while the cardinality of each set diverges, the ratio between the two remains finite.\footnote{Such treatment is exactly equivalent to what one does in information theory to extend Shannon's entropy to the continuous case.} We can redefine our cardinality as the ratio of our set and our reference set $I_0$. That is $n(I)=\hbar \delta q \delta k$ where $\hbar$ is chosen such that $n(I_0)=\hbar \delta q_0 \delta k_0=1$.\footnote{The choice of angular momentum for $\hbar$ derives from the relationship between $p$ and $dq/dt$, which will be derived later. The scale in classical mechanics is arbitrary: we have no physical reason to choose a reference label set over another. In quantum mechanics, instead, we will have a preference: the set that corresponds to a single quantized system.} In general, the label set is not rectangular in $\langle q,k \rangle$ and we have $n(I)=\int \hbar dq \wedge dk$

\begin{defn}\label{conjugateMomentum}
We define the \emph{conjugate momentum} as $p=\hbar k$.
\end{defn}

As $\hbar dq \wedge dk$ is truly fundamental, it is customary to use $dq \wedge (\hbar dk) = dq \wedge dp$. This way the area formed by generalized coordinate and conjugate momentum corresponds to the cardinality of states defined on such area.

\begin{cor}\label{continuousConjugateRelationships}
Each degree of freedom is identified by the pair of labels $\langle q,p \rangle \in QT^*$. $\theta_0 = p dq$ and $\omega = dq \wedge dp$ are invariant, $p$ is contravariant. The configuration state over a degree of freedom is $\mathbbm{c}=\int \rho(q,p) dq \wedge dp \mathbbm{s}_{qp}$, where $\rho(q,p)\equiv D(q,p) / \omega$ is the distribution density for each label.
\end{cor}

This restates all the previous findings in terms of conjugate momentum.

In the language of differential geometry, we recognise phase space as $Q$ extended by its cotangent space $T^*$, the function $\rho$, the one-form $\theta_0$ and the two-form $\omega$. They are truly fundamental objects as they are intimately linked to the way states are defined ($\rho$ is the distribution, $\theta_0 / \hbar$ is the cell width) and counted ($\omega$ is the cardinality of the labels) and they don't depend on coordinate choice.
